An Efficient Feedback Calibration Algorithm for Direct Imaging Radio Telescopes

### 2016-05-03 cal_demo/ms.tex
More comments on gridding, justifying the 2D FFT

Finally we assume the sources of the electric field are far away, allowing us to describe the 
electric fields propagating from sky sources as originating from an imaginary celestial sphere 
located a large distance from the ground. We express this intermediary electric field as an 
angular distribution, $\mathcal{E}(\hat{\mathbf{s}},f,t)$, where $\hat{\mathbf{s}}$ is the unit 
vector in the direction of a patch of the celestial sphere. We further assume the space between 
the celestial sphere and the ground is empty, allowing us to express the electric field incident 
on the ground as the integral of the propagated fields from the sphere.
\begin{equation}
E(\mathbf{r},f,t) = \int \mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega
\end{equation}  
By making the latter assumptions we have left treatment of the ionosphere and terrestrial 
emission such as local interference or thermal radiation from the ground for future work.

@@

The next step is to grid the calibrated electric field measurements to regularly spaced 
positions on the 2D plane of the ground, $\mathbf{r}_j$. 
MOFF uses the antenna patterns as the gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. 
Mathematically we can express the gridding operation as
\begin{equation}
E'_j=\sum_a W_a(\mathbf{r}_j-\mathbf{r}_a) E'_a(t)
\end{equation}
This 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric field measurements directly. This 
operation must be performed for every sample, before any averaging of the data.

After gridding, the imaging portion of the MOFF performs a 2D spatial FFT to form 
instantaneous electric field images of the sky. The pipeline up to this point can be summarized 
through the following equation.
\begin{equation}
\Ethetaest (\s_i,t) = \underbrace{\frac{1}{\Nant} \sum_i e^{2\pi i \ri \cdot \s_i}}_{\mathrm{2D\;FFT}} 
\underbrace{\sum_a \beamr_a(\ri - \ra)}_{\mathrm{Gridding}}
\underbrace{ \frac{g^T_a}{g'_a} \Er{a}^T(t)}_{\mathrm{Calibration}}
\end{equation}
The sum over antennas uses the voltage pattern to grid the calibrated electric fields onto 
regular gridpoint, $i$. The sum over $i$ denotes the 2D FFT to sky coordinates, resulting in an 
estimate for the instantaneous electric field image. We can simplify this expression by 
exchanging the sums to transform the beam term into sky coordinates.
\begin{align}\label{eq:epix}
\Ethetaest(\s_i,t) & = \frac{1}{\Nant} \sum_a \frac{g^T_a}{g'_a}\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra} \nonumber \\
  & \qquad\times \sum_i \beamr_a(\ri-\ra)e^{2\pi i \s_i \cdot (\ri-\ra)} \nonumber\\
& = \frac{1}{\Nant} \sum_a \frac{ g^T_a}{g'_a}\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra}\beamtheta_a(\s_i)
\end{align}
In this form we see that the effect of gridding with the antenna voltage pattern is to attenuate 
the image by a factor of the antenna sky response, $\beamtheta_a(\s_i)$.

>>>>

Finally we assume the sources of the electric field are far away, allowing us to describe the 
electric fields propagating from sky sources as originating from an imaginary celestial sphere 
located a large distance from the ground. We express this intermediary electric field as an 
angular distribution, $\mathcal{E}(\hat{\mathbf{s}},f,t)$, where $\hat{\mathbf{s}}$ is the unit 
vector in the direction of a patch of the celestial sphere. We further assume the space between 
the celestial sphere and the ground is empty, allowing us to express the electric field incident 
on the ground as a Fourier integral of the propagated fields across the hemisphere above the horizon.
\begin{equation}\label{eq:sky_propagated}
E(\mathbf{r},f,t) = \int \mathcal{E}(\hat{\mathbf{s}},f,t) e^{-2\pi i \mathbf{r} \cdot \hat{\mathbf{s}}}\, d\Omega
\end{equation}  
By making the latter assumptions we have left treatment of the ionosphere and terrestrial 
emission such as local interference or thermal radiation from the ground for future work.

@@

The next step is to grid the calibrated electric field measurements to a finites set of regularly spaced 
positions on the 2D plane of the ground, $\mathbf{r}_j$.
The extent of the grid is chosen to encompass the footprint of the antenna array.
MOFF uses the antenna aperture distributions as the gridding 
kernel according to the software holography/A-transpose technique \citep{mor09,bha08}. 
Mathematically we can express the gridding operation as
\begin{equation}
E'_j=\sum_a W_a(\mathbf{r}_j-\mathbf{r}_a) E'_a(t)
\end{equation}
This 
achieves an optimal map \citep{teg97b} and places the data on a regular grid. Unlike 
traditional correlators which grid visibilities, the MOFF grids the electric field measurements directly. This 
operation must be performed for every sample, before any averaging of the data.

After gridding, the imaging portion of the MOFF performs a 2D spatial FFT to form 
instantaneous electric field images of the sky at fixed locations $\s_k$.
The Fourier dual coordinates to $\mathbf{r}$ are $(l,m)$, where $l=\sin\theta\cos\phi$,
$m=\nobreak\sin\theta\sin\phi$, and $(\theta,\phi)$ are the zenith 
and azimuthal angles. The full unit vector towards the sky is 
$\s=(l,m,\sqrt{1-l^2-m^2})$.
The sample spacing and resolution of $\s_k$ are given 
by the choices of $\mathbf{r}_j$ grid parameters, according to the usual FFT relations.
The pipeline up to this point can be summarized 
through the following equation.
\begin{align}
\mathcal{E}'(\s_k,t) & = \underbrace{\frac{1}{\Nant} \sum_j e^{2\pi i \mathbf{r}_j \cdot \s_k}}_{\mathrm{2D\;FFT}} 
\underbrace{\sum_a W_a(\mathbf{r}_k - \ra)}_{\mathrm{Gridding}} \\
& \qquad \times \underbrace{ \frac{1}{g'_a} \left(g_a E_a^I(t)+n_a(t)\right)}_{\mathrm{Calibration}} \nonumber
\end{align}
The sum over antennas uses the voltage pattern to grid the calibrated electric fields onto 
regular gridpoint, $i$. The sum over $i$ denotes the 2D FFT to sky coordinates, resulting in an 
estimate for the instantaneous electric field image. We can simplify this expression by 
exchanging the sums to transform the beam term into sky coordinates.
\begin{align}\label{eq:epix}
\Ethetaest(\s_i,t) & = \frac{1}{\Nant} \sum_a \frac{g^T_a}{g'_a}\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra} \nonumber \\
  & \qquad\times \sum_i \beamr_a(\ri-\ra)e^{2\pi i \s_i \cdot (\ri-\ra)} \nonumber\\
& = \frac{1}{\Nant} \sum_a \frac{ g^T_a}{g'_a}\Er{a}^T(t) e^{2\pi i \s_i \cdot \ra}\beamtheta_a(\s_i)
\end{align}
In this form we see that the effect of gridding with the antenna voltage pattern is to attenuate 
the image by a factor of the antenna sky response, $\beamtheta_a(\s_i)$.
